```{r}

library("mice")
library("magrittr")

require(ggplot2)
```

```{r}
data(airquality)

colSums(is.na(airquality))

```

```{r}
md.pattern(airquality) # from mice package
```

```{r}
sum(complete.cases(airquality))
```

# Deletion methods

## Pairwise deletion example

```{r}
data <- airquality[, c("Ozone", "Solar.R", "Wind")]
sum(complete.cases(airquality))
```

Here two statistics are computed from the data: means and covariance

### Means with pairwise deletion

When taking the mean of each column, you can drop only the observations with missing values on that column (in complete case-analysis, even if the sample had values missing on other columns only, it would have been dropped).

```{r}

mu.listwise <- colMeans(na.omit(data))
mu.pairwise <- colMeans(data, na.rm = TRUE)
mu.difference <- mu.listwise-mu.pairwise
rbind(mu.listwise, mu.pairwise, mu.difference)
```

There is a difference between both methods, although not significant (Wilcoxon test)

```{r}
shapiro.test(mu.difference)
```

```{r}

wilcox.test(mu.listwise, mu.pairwise, paired = TRUE, alternative = "two.sided")
```

### Covariance with pairwise deletion

When computing covariance, a similar reasoning can be done: only drop observations if values on one of the two considered variables are missing. To do that in R you can give the argument `use="pairwise"`.

```{r}
cv <- cov(data, use = "pairwise")
cv
```

```{r}
library("lattice")
levelplot(cv, col.regions = topo.colors(100))
```

# Simple Imputation 

## Mean imputation

```{r}
imp <- mice(airquality, method = "mean", m = 1, maxit = 1)
imp$imp
xyplot(imp, Ozone ~ Solar.R)
```

```{r}
mu.mean <- colMeans(mice::complete(imp))
mu.mean
```

```{r}
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
summary(fit,fit.measures=TRUE, rsquare=TRUE, standardize=TRUE)
```

## Regression imputation

```{r}
imp <- mice(airquality, method = "norm.predict", seed = 1,
           m = 1, print = FALSE)
imp$imp
xyplot(imp, Ozone ~ Solar.R)
```

```{r}
mu.reg <- colMeans(mice::complete(imp))
mu.reg
```

```{r}
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
summary(fit,fit.measures=TRUE, rsquare=TRUE, standardize=TRUE)
```

## Stochastic non-Bayesian regression imputation

```{r}
imp <- mice(airquality, method = "norm.nob", m = 1, maxit = 1,
            seed = 1, print = FALSE)
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
xyplot(imp, Ozone ~ Solar.R)
xyplot(imp, Ozone ~ Wind)
```

```{r}
mu.sreg <- colMeans(mice::complete(imp))
mu.sreg
```

```{r}
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
summary(fit,fit.measures=TRUE, rsquare=TRUE, standardize=TRUE)
```

## Bayesian regression imputation with normality assumption

```{r}
imp <- mice(airquality, seed = 1, m = 1, method="norm", print = FALSE)
xyplot(imp, Ozone ~ Solar.R)
```

```{r}
mu.breg <- colMeans(mice::complete(imp))
mu.breg
```

```{r}
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
summary(fit)
```

# Multiple imputation

### Bayesian regression imputation with normality assumption

```{r}
imp <- mice(airquality, seed = 1, m = 20, method="norm", print = FALSE)
xyplot(imp, Ozone ~ Solar.R)
xyplot(imp, Ozone ~ Solar.R)
```

```{r}
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
summary(pool(fit))
```

### Parallel Mean Matching

```{r}
imp <- mice(airquality, seed = 1, m = 20, method="pmm", print = FALSE)
xyplot(imp, Ozone ~ Solar.R)
```

```{r}
fit <- with(imp, lm(Ozone ~ Wind + Temp + Solar.R))
summary(pool(fit))
```

# Full-information maximum likelihood

In this example we fit a linear model using full-information ML. This is another way of working around the missing data and use all the available data.

We use the `lavaan` package: [The lavaan Project (ugent.be)](https://lavaan.ugent.be/tutorial/est.html)

Notice the first argument `model`. The first line tells lavaan to predict ozon from wind + solar.R plus intercept. Second line tells lavaan to also fit variance for Ozone variable.

```{r}
library(lavaan)

fit<-sem(model="Ozone ~ 1 + Wind + Solar.R 
          Ozone ~~ Ozone",
          data=data,
          missing='ML',
          meanstructure=TRUE, 
          fixed.x=FALSE)
summary(fit, fit.measures=TRUE, rsquare=TRUE, standardize=TRUE)
```

## 
